Minimizing Time Required For Graduation

### 2010-12-12 paper/paper.tex
elminate some passive voice and weasel words; additional editing / improvements

%%% PACKAGES
\usepackage{booktabs} % for much better looking tables
\usepackage{array} % for better arrays (eg matrices) in maths
\usepackage{paralist} % very flexible & customisable lists (eg. enumerate/itemize, etc.)
\usepackage{epsfig}
\usepackage{verbatim} % adds environment for commenting out blocks of text & for better verbatim
\usepackage{subfig} % make it possible to include more than one captioned figure/table in a single float
\usepackage{amstext}
% These packages are all incorporated in the memoir class to one degree or another...

%%% HEADERS & FOOTERS
\usepackage{fancyhdr} % This should be set AFTER setting up the page geometry
\pagestyle{fancy} % options: empty , plain , fancy
\renewcommand{\headrulewidth}{0pt} % customise the layout...
\lhead{}\chead{}\rhead{}
\lfoot{}\cfoot{\thepage}\rfoot{}

@@

\title{Minimizing Time Required For Graduation}
\author{Ruyi Cai \and Matthew Fonda \and Michael Hughes}
%\date{} % Activate to display a given date or no date (if empty),
         % otherwise the current date is printed

@@

\abstract{The Bachelor's Degree is designed to take four years to complete;
however, many students fail to complete it in this time frame. One 
factor contributing to this is poor planning when registering for classes.
We present an algorithm to assist students in constructing a schedule 
that works towards completing a degree in the shortest amount of time 
possible. The algorithm presented uses heuristics to attempt to pick the best
class set each quarter, working towards a complete schedule quarterly by 
quarter.}
\pagebreak

\section{Introduction} \subsection{Background} Registration can be a stressful 
time for students at the University of Washington. Without proper planning, 
a student can run the risk of making poor registration choices, which can 
result in the student taking more quarters than otherwise necessary. However, 
choosing the right schedule is complex and there are multiple variables to 
consider. Picking classes involves the consideration of quarterly class 
offerings, class credits, prerequisites, time conflicts, and whether or not 
classes will fulfill future prerequisites. All of these factors can affect a 
student’s graduation time, especially if that class has limited offerings 
during a particular year. The University of Washington registration system 
already provides a tool to aid students: schedule finder. However, schedule 
finder only takes into account time conflicts per quarter, and not what 
classes are best for minimizing graduation time.

\subsection{Goal} Our goal is to provide an automated system that gives quarter
by quarter schedules up until graduates. As input our system will take in
the current set of classes a student has taken and the preferred credit hours
per quarter. Additional input will be the current and past university time
schedules. As output our system will provide a class schedule that minimizes the
number of quarters until graduation.

\subsection{Importance} The implementation of our system as part of the school’s
registration system will benefit the students by enabling them to map out their
quarterly class schedules until graduation. Students will be able to preview
schedules and the length of time until graduation.  This will lead to better
decision making for what majors to pursue, and given the student's current
state, which schedules are the most ideal in terms of saving time and resources.
This will also help students save money, as they will be in school for fewer
quarters. The university also benefits from the reduced labor cost and increased
efficiency of the advising offices.

\section{Problem Description} \label{pdesc} To make our examples concrete we will focus on the
classes offered within the Math department; however, our algorithm and model can
be extended to include other departments.

Every quarter the Mathematics department offers a selection of courses; not every
course is offered every quarter. Each course is offered during a certain set
time periods. There are inevitable time conflicts among courses that a student
could take because there are far more courses offered than time periods during
the regular five-day week. Graduating with a B.S. in Mathematics requires
specific classes in addition to elective requirements.
Additionally many classes have prerequisites which much be satisfied before they
can be taken. In the case of math courses, there can be multiple, exclusive
prerequisite sets that independently satisfy the requirements for a class.

Our goal is to determine a schedule of classes from some starting point to
graduation that given the above properties minimizes the total number of
quarters needed. Explicitly: \\ Minimize the total
number of quarters needed to graduate subject to: \begin{enumerate} \item Only
taking courses during quarters they are available \item Only taking a course
once \item Never exceeding a preset number of credit hours per quarter \item
Never taking more than one class at a given time \item Always satisfying
a course's prerequisites prior to taking that course \end{enumerate}

@@

\subsection{Classes and Quiz Sections} \label{classDefs} Another modeling
difficulty that we initially ignore was how to represent quiz sections for
a class. Each time a course is offered is may be offered as multiple classes.
For instance, during $F$ $MATH307$ is offered as 7 different sections all
teaching the same material during different time periods. Within a specific
quarter schedule line number, SLNs uniquely identify courses allowing us to
differentiate between classes with the same name. Continuing the above example,
the SLNs of the 7 $MATH307$ instances may be 15432, 15433, 15434...

Our peer review team commented that we had not accounted for courses that
required concurrent quiz sections. First we note that the quiz sections for
a class never occur during the same time periods as the class itself (otherwise
students would not be able to attend the quiz section). Quiz sections also have
unique SLNs, to include a required quiz section we include the quiz section SLN
with the SLN for the associated class. Finally, courses such at $MATH125$ can
have multiple sections, each with several quiz sections of which only one need
be taken. We found the easiest way to represent this situation was to iterate
over the quiz sections for each class section and generate a pseudo class for
each pair of class section and quiz section that occurred during the time periods
for both quiz and main sections. Fig. \ref{quizzes} contains one section
$MATH125$ being split across 3 quiz sections.
\begin{figure} [ht] 
    \begin{center}
        \includegraphics[scale=0.4]{quiz_sections} 
    \end{center}
    \caption{Splitting up MATH125} 
    \label{quizzes} 
\end{figure}

\subsection{Definitions} We will now formalize some terms used in our paper.
\begin{itemize} \item A {\it course} refers to any course offered by the
University of Washington. A course is identified by its department name and
number, for example {\it MATH309}. Each course is also worth a specific amount
of credits. \item We refer to a specific instance of a course as a {\it class}.
Each class has associated with it the time of the day at which it meets, the
quarter it is being offered, and the section number of the offering.  Throughout
this paper, we will refer to quarters by their time schedule abbreviates: {\it
F} for Fall Quarter, {\it W} for Winter Quarter, {\it Sp} for Spring Quarter,
and {\it S} for Summer Quarter. \end{itemize}

Additionally: \begin{itemize} \item Let $S$ be the set of courses required to
graduate. $S$ can be thought of as our goal state, and can also contain elective
courses a student simply wishes to take.  \item Let $T$ be the set of courses
a given student has already completed. If no courses have been taken, this set
may be empty.  \item Let $U$ be the set of all courses a student could possibly
ever take.  \item Let $R$ be a set of rules specifying course prerequisites.
\item Let $l$ be the maximum number of credits a student wishes to take during
any given quarter.  \item Let $O$ be the set of all 50-minute time blocks
between 8:30am and 7:30pm from Monday through Friday. \item Let $L_i$ be the set
of courses a student can take in quarter i. \end{itemize}

\subsection{Representing Prerequisites} Initially, we hoped to represent course
prerequisites using a directed acyclic graph. Each node was a course, and edges
existed between nodes $i$ and $j$ if and only if $i$ was a prerequisite for $j$.
The DAG relationship works well when courses have single requirements. It
becomes difficult, however, to track whether requirements are fulfilled with courses
that have multiple prerequisites, or worse courses that have alternative
prerequisites. We decided that it would be feasible to model prerequisites
using a graph.
\begin{figure} [ht] 
    \begin{center}
        \includegraphics[scale=0.4]{prereq_logic_example} 
    \end{center}
    \caption{Evaluation of a rule} 
    \label{logic_ex} 
\end{figure}

To represent alternative and multiple prerequisites we introduced a set of logic
rules for each course. The prerequisites for each course are expressed as
propositional logic statements stored in the set $R$. The literal propositions
in each rule are related to whether or not a student has taken a particular
course. For example, the rule associated with $MATH327$ in fig~\ref{logic_ex} is
$$ \text{{\it MATH327: (MATH126 AND MATH300) OR MATH136}}.$$

To determine if a given student is eligible to register for a course offering,
c we can evaluate the rule for c. Each literal proposition in the rule takes on
the value true if a student has already taken the course, and false if he or she
hasn't.  If the rule evaluates to true, then the student is eligible to take the
course; otherwise. For example, consider Figure~\ref{logic_ex}, here a student
has previously taken courses $MATH126$ and $MATH300$. Since $MATH126$ and
$MATH300$ have been taken, then those literals in the rule for $MATH327$
evaluate to true and therefore the entire rule is true.  Since the entire rule
is true, the student can take $MATH327$.

\section{Tractability} With a static schedule it would be
possible to plan an entire set of quarter by quarter schedules from start to
finish. Unfortunately we have neither the computational resources nor the luxury
of a static schedule. Some classes are offered every quarter others such as
senior electives are not. Furthermore for some classes it cannot be predicted
more than one quarter in advance when a senior elective will be offered. Even
given perfect information, where we know the possible class schedules for every
quarter we face another problem; finding the best possible schedule reduces to
enumerating possible class schedules for every quarter until the graduation
requirements are satisfied.  Given that we cannot predict in advance the
conflicts between classes, we cannot even determine the branching factor of the
enumeration tree nor its depth. Since we can only determine one quarter in
advance, this suggests approaching the problem by using heuristics to pick the
best possible classes each quarter.

Instead of trying to solve the complete problem we chose to use the "rolling
horizon approach" used by Xiangtong et al \cite{xiangton:informs} when
scheduling classes for airline pilots. Xiangtong et al. also could not develop
an optimal class schedule in one pass, instead they picked optimal classes based
using a heuristic over a well demarcated time period. Once they had a solution
to the smaller time period, they increase the size of the time period, or rolled
the horizon, and used the previously optimal solution to help determine
a solution to the new larger problem. The team's algorithm continued rolling the
time horizon on the problem until they had a solution to the entire problem.

\section{Picking Classes Per Quarter} Since determining an entire schedule at
once is intractable we adapted the rolling horizons approach which uses
a heuristic to pick the best possible schedule for a quarter. With a good
heuristic we will be able to choose classes that lead to a minimized graduation
time. Since we cannot analytically determine graduation time based on classes
chosen in a single quarter, our heuristic tries to give high value to classes
that would delay graduation is taken later rather than sooner.

@@

\subsection{Determining Available Course Offerings} We need to select the
classes we can currently take and then build conflict graphs for classes based
on the times they are offered. The set $L_i$ is built in this part of the
algorithm.

\subsubsection{Procedure} We present the procedure for populating the set $L$ as
follows: \begin{enumerate} \item For each rule $r \in R$ we iterate over each
proposition $p$ in the rule. Recall that $p$ represents a course, for example
$MATH125$. We then check if $p \in T$.  If $p \in T$ then we say $p$ evaluates
to true; otherwise, $p$ evaluates to false.  Using rules of logic we evaluate
$r$, if $r$ true, then we add the course associated with it to $L$.  Once this
is complete, $L$ will consist of every course that could be taken based solely
on prerequisites that we have fulfilled.  \item We now prune $L$ by removing all
courses that are not offered in the current quarter. $L$ now only contain
classes that we can take given the current quarter and the classes we have
already taken.  \item We now construct a conflict graph, $G_{conflict}$, where
every course offering in $L$ is a node, and there is an edge between two nodes
$i$ and $j$ if and only if the time of course offering $i$ overlaps the time of
course offering $j$. An example conflict graph for one time period is given in
Fig. \ref{conflictg}. \end{enumerate} 
\begin{figure} [ht] 
    \begin{center}
        \includegraphics[scale=0.35]{conflict_graph_ex} 
    \end{center} 
    \caption{A Conflict Graph} 
    \label{conflictg} 
\end{figure}

\subsubsection{Analysis} The selection procedure involves a lot of set queries
and iteration, but in asymptotic running time the algorithm is relatively quick.
Let $n$ be the total number of classes ever available to take, or in other
words, $n = \|U\|$. In the worst case, $P$ has a complex rule for all $n$
courses, and checking each rule in $P$ is linear in the number of propositions.
However, we know {\it a priori} from analysis of our data set that the largest
rule has twelve propositions, so we can bound the rule check time by a constant,
$C$, and therefore step one can be done in $O(n)$ time.

Continuing to step two, the worst case is when $L$ = {all classes offered during
the quarter}.  Checking whether a course is offered in a given quarter can be
done in constant time, and we have $n$ courses to check, therefore step two can
also be done in $O(n)$ time.

@@

\subsection{Selection of Quarterly Classes} To select classes we give each class
a value and weight. The primary problem is the the classes contained in $L$
almost certainly have time conflicts, likely between courses that need to be
taken.

\subsubsection{Heuristic Value Function} In order to determine which classes
should be picked in a given schedule, we use a heuristic function which assigns
each class a value. The higher the value of this function for a class, the more
likely that class will be placed in the schedule.  Recalling that our goal is to
minimize the total number of quarters taken (and implicitly, maximize the total
number of courses taken per quarter).  Our heuristic function should give higher
value to class that maximize the number of classes taken this quarter and in
future quarters.  One component to the heuristic function then must be the
number of dependent courses a given course offering has. If a course satisfies
many prerequisites, then we know that there are many courses that we cannot
complete without it, so we want our heuristic function to favor it.

@@

Where $c$ is a given course offering, $q$ is the current quarter, and $\alpha$,
$\beta$, and $\gamma$ are real numbers, giving a weight to piece of the
heuristic. The scalars are determined experimentally.

\subsubsection{Selection Procedure} At this point in the algorithm we know which
courses are available to a student and their relative values. We present two
ways of approaching quarterly class choice.  We will initially encode the class
selection as a binary integer programming problem in spirit of the methods
presented by Dinkel et al \cite{dinkel:scheduling}. and Pritsker et al
\cite{prisker:informs}. A second approach developed by Yamada et al
\cite{yamada:heuristic} and Pferschy et al \cite{pferschy:kcg} is presented for
dealing with very large class choices where a binary LP would be difficult
solve.

@@

The linear program can be interpreted as follows: \begin{itemize} \item
Eqn.~\ref{qtrval}: The value of a quarter is the total number of courses taken
that quarter. We want to maximize this value, to maximize the number of high
value classes we take each quarter. \item Eqn.~\ref{weightlim}: In each quarter,
the total number of credits cannot exceed the specified limit. $Weight(c_i)$
represents a short hand way of indicating the number of credits hours a class
takes. \item Eqn.~\ref{timecon}: Only one course can be taken in one particular
time slot, this is where $G_{conflict}$ is used. Without the prebuilt conflict
graph it would be difficult to build the set of constraints based on equation
number~ref{timecon}. \item Eqn.~\ref{coursecon}: Previously in
section~\ref{classDefs} we noted that in many cases a course is offered in
multiple sections during the same quarter. A student can only take one offering
of a course, this constraint eqn must be built for every class-quiz section
combination which represent that same course.\end{itemize}

@@

\subsection{Alternative Formulation} As noted in the reference literature this
is actually a NP-hard problem as described \cite{pferschy:kcg}.  For versions of
this problem in the low hundreds of variables, however it is still possible to
solve using modern LP solving software \cite{yamada:heuristic}. In our case,
just solving for the optimal classes for a single major fits well into the scope
of a few hundred (or even just tens of) variables. For multiple majors or to
consider class selection for distribution and elective requirements, however the
number of classes balloons significantly and the KCG problem becomes much
harder. For this version of the problem we suggest the techniques used by Yamada
et al \cite{yamada:heuristic}, which we briefly outline below.

@@

In Yamada et al’s \cite{yamada:heuristic} numerical runtime analysis they found
this algorithm ran very favorably against solving the same problem using popular
linear programming solvers. There is one problem that we should briefly address,
because the number of periods of available classes per week is relatively small
the conflict rate for a large number of class is relatively higher than Yamada
et al’s 2\% maximal conflict rate for their experiments. Our data has
a significant mitigating factor though, knapsack problems in general are
exponential in terms of the number of choices and the total weight capacity.
Realistically the total weight capacity is going to be 20 credits or less which
is between 1\% and 2\% of the figures used by Yamada \cite{yamada:heuristic} and
also our total number of choices is in many cases (particular just the classes
in a particular major) less than 200. Giving the size and type of data in our
set the higher conflict rate is less of a concern and Yamada et al’s algorithm
should still be efficient.

>>>>

%%% PACKAGES
\usepackage{booktabs} % for much better looking tables
\usepackage{array} % for better arrays (eg matrices) in maths
\usepackage{paralist} % flexible & customisable lists (eg. enumerate/itemize, etc.)
\usepackage{epsfig}
\usepackage{verbatim} % adds environment for commenting out blocks of text & for better verbatim
\usepackage{subfig} % make it possible to include more than one captioned figure/table in a single float
\usepackage{amstext}
% These packages are all incorporated in the memoir class to one degree or another...

%%% HEADERS & FOOTERS
\usepackage{fancyhdr} % Set AFTER setting up the page geometry
\pagestyle{fancy} % options: empty , plain , fancy
\renewcommand{\headrulewidth}{0pt} % customise the layout...
\lhead{}\chead{}\rhead{}
\lfoot{}\cfoot{\thepage}\rfoot{}

@@

\title{Minimizing Time Required For Graduation}
\author{Ruyi Cai \and Matthew Fonda \and Michael Hughes}

@@

\abstract{The Bachelor's Degree takes four years to complete;
however, students often fail to complete it in this time frame. One
factor contributing to this is poor planning when registering for classes.
We present an algorithm to assist students in constructing a schedule 
that works towards completing a degree in the shortest amount of time 
possible. The algorithm presented uses heuristics to attempt to pick the best
class set each quarter, working towards a complete schedule quarterly by 
quarter.}
\pagebreak

\section{Introduction} \subsection{Background} Registration can be a stressful 
time for students at the University of Washington. Without proper planning, 
a student runs the risk of making poor registration choices resulting
in the student taking more quarters than otherwise necessary.
Choosing the right schedule is complex and there are multiple variables to
consider. Picking classes involves the consideration of quarterly class 
offerings, class credits, prerequisites, time conflicts, and whether or not 
classes will fulfill future prerequisites. All of these factors affect the amount
of time required for a student to graduate. The University of Washington registration
system already provides a tool to aid students: schedule finder. However, schedule
finder only takes into account time conflicts per quarter, and not what classes are
best for minimizing graduation time.

\subsection{Goal} Our goal is to provide an automated system that generates quarterly
schedules until the point of graduation. As input our system will take in
the current set of classes a student has taken, the preferred credit hours
per quarter of a student, and current and past university time schedules.
Our system will output a class schedule that minimizes the
number of quarters required until graduation.

\subsection{Importance} The implementation of our system as part of the school’s
registration system will benefit students by enabling them to map out their
quarterly class schedules until graduation. Students will be able to preview
schedules and the length of time until graduation.  This will lead to better
decision making for what majors to pursue, and given the student's current
state, which schedules are the most ideal in terms of saving time and resources.
This will also help students save money, as they will be in school for fewer
quarters. The university also benefits from the reduced labor cost and increased
efficiency of advising offices.

\section{Problem Description} \label{pdesc} To make our examples concrete we will focus on the
classes offered within the Math department; however, our algorithm also works with for other departments.

Every quarter the Mathematics department offers a selection of courses. This offering is a
subset of all Mathematics courses, and varies from quarter to quarter.
Each course is available within a specific set of
time periods. There are inevitable time conflicts among courses that a student
could take because there are far more courses offered than there are time periods during
the regular five-day week. Graduating with a B.S. in Mathematics requires
specific classes in addition to elective requirements.
Additionally most courses require the completion of prerequisite courses.
There can be multiple, disjoint prerequisite sets that independently
satisfy the requirements for a class.

Our goal is to determine a schedule of classes from some starting point to
graduation that given the above properties, minimizes the total number of
quarters required to graduate. Explicitly: \\ Minimize the total
number of quarters needed to graduate subject to: \begin{enumerate} \item Only
taking courses during quarters they are available \item Only taking a course
once \item Never exceeding a preset number of credit hours per quarter \item
Never taking more than one class at a given time \item Always satisfying
a course's prerequisites prior to taking that course \end{enumerate}

@@

\subsection{Classes and Quiz Sections} \label{classDefs} A modeling
difficulty that we initially ignore was accounting for quiz sections and
for multiple offering of the same course within one quarter.
For instance, the Fall Quarter time schedule includes seven different
offerings of $MATH307$. A schedule line number (SLN)
uniquely identifies each class, allowing us to differentiate between classes
with the same title. Continuing the above example,
the SLNs of the seven $MATH307$ offerings may be 15432, 15433, 15434...

Our peer review team commented that we did not account for courses that
required concurrent quiz sections. First we note that the quiz sections for
a class never occur during the same time periods as the class itself (otherwise
students would not be able to attend the quiz section). Quiz sections also have
unique SLNs, to include a required quiz section we include the quiz section SLN
with the SLN for the associated class. Finally, courses can have multiple sections
and multiple quiz sections. Additionally, we have the restriction that students
cannot concurrently enroll in more than one section of a given course. We found that
the easiest way of representing this situation is generating a pseudo-class for each
possible class and quiz section pair.
Fig. \ref{quizzes} illustrates splitting a $MATH125$ section into three pseudo-classes.
\begin{figure} [ht] 
    \begin{center}
        \includegraphics[scale=0.4]{quiz_sections} 
    \end{center}
    \caption{Splitting up MATH125} 
    \label{quizzes} 
\end{figure}

\subsection{Definitions} We now formalize some terms used in our paper.
\begin{itemize} \item A {\it course} refers to any course offered by the
University of Washington. A course is uniquely identified by its department name and
number, for example {\it MATH309}. Each course is worth a specific amount
of credits. \item We refer to a specific instance of a course as a {\it class}.
Each class meets at a specific time of day during a specific quarter. Additionally,
each class has a unique SLN. Throughout
this paper, we refer to quarters by their time schedule abbreviates: {\it
F} for Fall Quarter, {\it W} for Winter Quarter, {\it Sp} for Spring Quarter,
and {\it S} for Summer Quarter. \end{itemize}

Additionally: \begin{itemize} \item Let $S$ be the set of all courses required for
a student to graduate, additionally including any electives a student wishes to take.
$S$ is our goal state.  \item Let $T$ be the set of courses
a given student has already completed.
\item Let $R$ be a set of rules specifying course prerequisites.
\item Let $l$ be the maximum number of credits a student wishes to take during
any given quarter.  \item Let $O$ be the set of all 50-minute time blocks
between 8:30am and 7:30pm from Monday through Friday. \item Let $L_i$ be the set
of courses a student can take in quarter i. \end{itemize}

\subsection{Representing Prerequisites} Initially, we represented course
prerequisites using a directed acyclic graph. Each node represented a course, and edges
existed between nodes $i$ and $j$ if and only if $i$ was a prerequisite for $j$.
This model works well if each course has only one prerequisite. However,
this model fails to represent courses that have multiple prerequisites
or courses that have alternative prerequisites. This is the case for some courses.
We therefore concluded that it is infeasible to model prerequisites using a graph.
\begin{figure} [ht] 
    \begin{center}
        \includegraphics[scale=0.4]{prereq_logic_example} 
    \end{center}
    \caption{Evaluation of a rule} 
    \label{logic_ex} 
\end{figure}

To represent alternative and multiple prerequisites we introduced a set of logic
rules for each course. The set $R$ consists of a prepositional logic statements representing
the prerequisites of each course. The literals in each statement are a course.
For example, the rule associated with $MATH327$ in Figure~\ref{logic_ex} is
$$ \text{{\it MATH327: (MATH126 AND MATH300) OR MATH136}}.$$

To determine if a given student is eligible to register for a given class $c$,
we evaluate the rule for $c$. Each literal in the statement takes on
the value true if a student has already taken the course, and false if the student
hasn't. If the statement evaluates to true, then the student is eligible to take the
course; otherwise, the student is not. For example, consider Figure~\ref{logic_ex}. Here, a student
has previously taken courses $MATH126$ and $MATH300$. Since the student completed both $MATH126$ and
$MATH300$, these literals in the rule for $MATH327$ evaluate to true,
and therefore the entire rule evaluates true. The student is therefore eligible to take $MATH327$.

\section{Tractability} With a static schedule it would be
possible to plan an entire set of quarter by quarter schedules from start to
finish. Unfortunately we have neither the computational resources nor the luxury
of a static schedule. Furthermore, the next offering of a given course is unpredictable.
Even with perfect information, where we know the possible class schedules for every
quarter we face another problem; finding the best possible schedule reduces to
enumerating possible class schedules for every quarter until the fulfillment of graduation
requirements. Given that we cannot predict in advance the
conflicts between classes, we cannot even determine the branching factor of the
enumeration tree nor its depth. Since we can only determine one quarter in
advance, this suggests approaching the problem by using heuristics to pick the
best possible classes each quarter.

Instead of trying to solve the complete problem we chose to use the "rolling
horizon approach" used by Xiangtong et al \cite{xiangton:informs} when
scheduling classes for airline pilots. Xiangtong et al. also could not develop
an optimal class schedule in one pass, instead they picked optimal classes
using a heuristic over a well demarcated time period. Once they had a solution
to the smaller time period, they increased the size of the time period, or rolled
the horizon, and used the previously optimal solution to help determine
a solution to the new larger problem. The team's algorithm continued rolling the
time horizon on the problem until they had a solution to the entire problem.

\section{Picking Classes Per Quarter} Since determining an entire schedule at
once is intractable, we adapted the rolling horizons approach which uses
a heuristic to pick the best possible schedule for a quarter. With a good
heuristic we will be able to choose classes that lead to a minimized graduation
time. Since we cannot analytically determine graduation time based on classes
chosen in a single quarter, our heuristic gives higher value to classes
that delay graduation the longest.

@@

\subsection{Determining Available Course Offerings} We need to select the
classes we can currently take and then build time conflict graphs.
We construct the set $L_i$ during this phase of the algorithm.

\subsubsection{Procedure} We present the procedure for populating the set $L$ as
follows: \begin{enumerate} \item For each rule $r \in R$ we iterate over each
literal $p$ in the rule. Recall that $p$ represents a course, for example
$MATH125$. We then check if $p \in T$.  If $p \in T$, we say $p$ evaluates
to true; otherwise, $p$ evaluates to false.  Using rules of logic we evaluate
$r$. If $r$ is true, we add the course associated with it to $L$.  Once this
is complete, $L$ will consist of every course a student is eligible to take, based solely
on satisfied prerequisites.  \item We now prune $L$ by removing all
courses that are not offered in the current quarter. $L$ now contains
only classes that we can take given the current quarter and the classes we have
already taken.  \item We now construct a conflict graph, $G_{conflict}$, where
every course offering in $L$ is a node, and there is an edge between two nodes
$i$ and $j$ if and only if the time of course offering $i$ overlaps the time of
course offering $j$. Figure~\ref{conflictg} shows an example conflict graph for
one time period. \end{enumerate}
\begin{figure} [ht] 
    \begin{center}
        \includegraphics[scale=0.35]{conflict_graph_ex} 
    \end{center} 
    \caption{A Conflict Graph} 
    \label{conflictg} 
\end{figure}

\subsubsection{Analysis} The selection procedure involves a lot of set queries
and iteration, but in asymptotic running time the algorithm is efficient.
Let $n$ be the total number of classes ever available to take.
In the worst case, $P$ has a complex rule for all $n$
courses, and checking each rule in $P$ is linear in the number of propositions.
However, we know {\it a priori} from analysis of our data set that the largest
rule has twelve propositions, so we can bound the rule check time by a constant,
$C$, and therefore step one runs in $O(n)$ time.

Continuing to step two, the worst case is when $L$ = {all classes offered during
the quarter}.  Checking whether a course is offered in a given quarter takes
constant time, and we have $n$ courses to check, therefore step two also
runs in $O(n)$ time.

@@

\subsection{Selection of Quarterly Classes} To select classes we give each class
a value and weight. The primary problem is the classes contained in $L$
almost certainly have time conflicts, likely between courses that need to be
taken.

\subsubsection{Heuristic Value Function} To pick the classes taken in a given quarter,
we use a heuristic function which assigns
each class a value. When constructing the schedule, we pick classes with the highest heuristic value.
Recall that our goal is to
minimize the total number of quarters taken (and implicitly, maximize the total
number of courses taken per quarter). Therefore, our heuristic function gives higher
value to class that maximize the number of classes taken in a given quarter and in
future quarters.  One component to the heuristic function then must be the
number of dependent courses a given course offering has. If a course satisfies
prerequisites, then there are courses that we cannot
complete without it. Therefore, we want our heuristic function to favor it.

@@

Where $c$ is a given course offering, $q$ is the current quarter, and $\alpha$,
$\beta$, and $\gamma$ are real numbers, giving a weight to piece of the
heuristic. We determine these scalars experimentally.

\subsubsection{Selection Procedure} At this point in the algorithm we know which
classes are available to a student and the values of the classes. We present two
ways of approaching quarterly class choice.  We will initially encode the class
selection as a binary integer programming problem in spirit of the methods
presented by Dinkel et al \cite{dinkel:scheduling}. and Pritsker et al
\cite{prisker:informs}. We then present a second approach developed by Yamada et al
\cite{yamada:heuristic} and Pferschy et al \cite{pferschy:kcg} for
dealing with large class choices where a binary linear program is difficult to solve.

@@

We interpret the linear program as follows: \begin{itemize} \item
Eqn.~\ref{qtrval}: The value of a quarter is the total number of courses taken
that quarter. We want to maximize this value, to maximize the number of high
value classes we take each quarter. \item Eqn.~\ref{weightlim}: In each quarter,
the total number of credits cannot exceed the specified limit. $Weight(c_i)$
represents a short hand way of indicating the number of credits hours a class
takes. \item Eqn.~\ref{timecon}: A student can take only one class in one particular
time slot. Recall that we use $G_{conflict}$ to determine time conflicts.
\item Eqn.~\ref{coursecon}: Previously in
section~\ref{classDefs} we noted that it is possible for a course to have multiple
offerings during the same quarter. A student can only take one offering
of a course during a quarter. We build this constraint for every class-quiz section
combination for a given course.\end{itemize}

@@

\subsection{Alternative Formulation} As noted in the reference literature this
is actually a NP-hard problem as described \cite{pferschy:kcg}.  For versions of
this problem in the low hundreds of variables, however it is still possible to
solve using modern LP solving software \cite{yamada:heuristic}. In our case,
just solving for the optimal classes for a single major is on the order of
hundreds of decision variables. For multiple majors or to
consider class selection for distribution and elective requirements, however the
number of classes balloons and the KCG problem becomes infeasible.
For this version of the problem we suggest the techniques used by Yamada
et al \cite{yamada:heuristic}, which we briefly outline below.

@@

In Yamada et al’s \cite{yamada:heuristic} numerical runtime analysis they found
this algorithm ran favorably against solving the same problem using popular
linear programming solvers. There is one problem that we should briefly address,
because the number of periods of available classes per week is small,
the conflict rate for a large number of class is higher than Yamada
et al’s 2\% maximal conflict rate for their experiments. Our data has
a significant mitigating factor though, knapsack problems in general are
exponential in terms of the number of choices and the total weight capacity.
Realistically the total weight capacity is going to be 20 credits or less which
is between 1\% and 2\% of the figures used by Yamada \cite{yamada:heuristic} and
also our total number of choices is in some cases (particular just the classes
in a particular major) less than 200. Giving the size and type of data in our
set the higher conflict rate is less of a concern and Yamada et al’s algorithm
should still be efficient.
